---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
First we load the raw activity data from the `activity.zip` file.
```{r echo=TRUE}
zipped_activity <- unz("activity.zip", filename = "activity.csv")
activity_data <- read.csv(zipped_activity, 
                          colClasses = c("numeric", "POSIXct", "numeric"))
```

## What is mean total number of steps taken per day?

First we summarize the number of steps observed for each day:
```{r echo=TRUE}
library(dplyr, warn.conflicts = FALSE)

daily_step_totals <- group_by(activity_data, date) %>% 
    summarize(total_steps = sum(steps, na.rm = TRUE))
```

The histogram below shows the distribution of daily totals:
```{r echo=TRUE}
hist(daily_step_totals$total_steps,  
     main = "Total Steps by Day", 
     xlab = "Total Steps")
```

The mean of this data is `r mean(daily_step_totals$total_steps)` and the median is
`r median(daily_step_totals$total_steps)`.

## What is the average daily activity pattern?
First, we get the average number of steps for each interval period across
all observed days:
```{r echo=TRUE}
interval_step_averages <- 
    group_by(activity_data, interval) %>% 
    summarize(mean_steps = mean(steps, na.rm = TRUE))
```

Now we can graph this data set:
```{r echo=TRUE}
plot(interval_step_averages$interval, 
     interval_step_averages$mean_steps, 
     type="l",
     xlab = "Interval", 
     ylab = "Mean Steps")
```

```{r echo=TRUE}
highest_interval <- filter(interval_step_averages, mean_steps == max(mean_steps))
```
The interval with the highest mean number of steps is `r highest_interval[[1,1]]`.

## Imputing missing values
```{r echo=TRUE}
num_missing <- nrow(filter(activity_data, is.na(steps)))
```
The data-set is missing several values for certain intervals. A total of
`r num_missing` observations have no recorded steps activity.

For each observation with a missing `steps` value, we replace it with the
_mean_ of observed steps for the same interval.
```{r echo=TRUE}
interval_means <- activity_data %>% 
    group_by(interval) %>% 
    summarize(mean_steps = mean(steps, na.rm = TRUE))

filled_activity_data <- 
    transform(activity_data, 
              steps = ifelse(is.na(steps), 
                             interval_means[interval_means$interval == interval]$mean_steps, 
                             steps))
```

We can look at the distribution of the imputed values with this histogram:
```{r echo=TRUE}
hist(filled_activity_data$steps,  
     main = "Total Steps by Day (Imputed)", 
     xlab = "Total Steps")
```

The _mean_ number of steps is `r mean(filled_activity_data$steps)` and the
_median_ number of steps is `r median(filled_activity_data$steps)`. When we 
fill in the missing observations the distribution of values skews much 
closer to zero. This is due to the prevlance of observations with 0 as the
observed value for the `steps` variable.

## Are there differences in activity patterns between weekdays and weekends?
To determine the differences in activity based the day of the week, we first
calculate a new factor variable in the `filled_activty_data` data.frame called
`day_type` which is a factor with two possible values: "weekday" and "weekend".

```{r echo=TRUE}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
filled_activity_data <- 
    transform(filled_activity_data,
              day_type = as.factor(ifelse(weekdays(date) %in% weekdays,
                                          "weekday",
                                          "weekend")))
```

Next, we'll break the data apart by `day_type` and compute the mean for
each `interval`:
```{r echo=TRUE}
weekday_activity <- 
    filter(filled_activity_data, day_type == "weekday") %>%
    group_by(interval) %>%
    summarize(mean_steps = mean(steps))

weekend_activity <- 
    filter(filled_activity_data, day_type == "weekend") %>%
    group_by(interval) %>%
    summarize(mean_steps = mean(steps))
```

Now we can compare these two sets with a pair of simple line-plots:
```{r echo=TRUE, fig.height=7, fig.width=7}
par(mfrow = c(2, 1), mar = c(4, 4, 4, 2), oma = c(1, 1, 1, 1))
with(weekday_activity, 
     plot(interval, mean_steps, type="l", cex=0.5,
          main="Weekdays", xlab="Interval", ylab="Steps"))

with(weekend_activity, 
     plot(interval, mean_steps, type="l", cex=0.5, 
          main="Weekends", xlab="Interval",  ylab="Steps"))    
```
